Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-12 treestats_paper.tex
Kreitman 1983

It was once a major undertaking to collect data sufficient to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate $F_{ST}$ between two populations using a dozen allozymes.
\jk{Citation?}
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread},
or between individuals rather than populations \citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}.
Computation is beginning to be a major problem:
genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
for instance, might have 500K individuals genotyped at 10M loci,
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Furthermore, even larger datasets are being collected,
and millions of whole genomes will soon be available. In this case the
genotype matrix simply cannot fit into memory, and complex distributed
computations are required to calculate population genetic statistics.

>>>>

It was once a major undertaking to collect data sufficient to estimate a single summary
of the many genetic relationships between the individuals of a sample,
e.g., to estimate levels of sequence diversity from a few dozen alleles at an allozyme locus
\citep{Kreitman1983-xr}.
Today's vast quantity of whole-genome sequence
makes it possible to confidently estimate many more properties of genealogical relationships
in local regions of genomes \citep[e.g,][]{booker2018understanding,haenel2018metaanalysis,stankowski2019widespread},
or between individuals rather than populations \citep[e.g.,][]{browning2010highresolution,speed2014relatedness,aguillon2017deconstructing}.
Computation is beginning to be a major problem:
genetic data on the scale of the UK Biobank~\citep{bycroft2018genome},
for instance, might have 500K individuals genotyped at 10M loci,
resulting in a genotype matrix of $5 \times 10^{12}$ entries.
Computing a single summary from this matrix is still relatively quick,
once loaded into memory,
but for many applications we are now interested in computing a large number of statistics,
or iteratively refining a single computation.
Furthermore, even larger datasets are being collected,
and millions of whole genomes will soon be available. In this case the
genotype matrix simply cannot fit into memory, and complex distributed
computations are required to calculate population genetic statistics.
